---
title: "ENVS193DS_homework-03"
author: "Maya Bernstein"
format: html
html: 
    toc: true # includes a table of contents in rendered html format
execute: # for all code chunks
    warning: FALSE # do not display warnings
    message: FALSE # do not display messages
---

## Link to Repository *add link*

```{r}
# general use
library(tidyverse)
library(readxl)
library(here)
library(janitor)

# visualizing pairs
library(GGally)

# model selection
library(MuMIn)

# model predictions
library(ggeffects)

# model tables
library(gtsummary)
library(flextable)
library(modelsummary)

drought_exp <- read_xlsx(path = here("data", 
                                     "Valliere_etal_EcoApps_Data.xlsx"),
                         sheet = "First Harvest")

# quick look at data 
str(drought_exp)
class(drought_exp)
```

```{r}
# cleaning
drought_exp_clean <- drought_exp %>% 
  clean_names() %>% # nicer column names
  mutate(species_name = case_when( # adding column with species scientific names
    species == "ENCCAL" ~ "Encelia californica", # bush sunflower
    species == "ESCCAL" ~ "Eschscholzia californica", # California poppy
    species == "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    species == "GRICAM" ~ "Grindelia camporum", # great valley gumweed
    species == "SALLEU" ~ "Salvia leucophylla", # Purple sage
    species == "STIPUL" ~ "Nasella pulchra", # Purple needlegrass
    species == "LOTSCO" ~ "Acmispon glaber" # deerweed
  )) %>% 
  relocate(species_name, .after = species) %>% # moving species_name column after species
  mutate(water_treatment = case_when( # adding column with full treatment names
    water == "WW" ~ "Well watered",
    water == "DS" ~ "Drought stressed"
  )) %>% 
  relocate(water_treatment, .after = water) # moving water_treatment column after water
```

# Question 1

## **Creating models**

#### Model 0 - null

```{r}
#null model
model0 <- lm(total_g ~ 1, # formula
             data = drought_exp_clean) # data frame
```

#### Model 1 - total biomass as a function of SLA, water treatment, and species

```{r}
# saturated model
model1 <- lm(total_g ~ sla + water_treatment + species_name, #formula
             data = drought_exp_clean) #dataframe
```

#### Model 2 - total biomass as a function of SLA and water treatment

```{r}
model2 <- lm(total_g ~ sla + water_treatment, #formula w correct predictors
             data = drought_exp_clean) #dataframe
plot(model2)
```

#### Model 3 - total biomass as a function of SLA and species

```{r}
model3 <- lm(total_g ~ sla + species_name, #formula
             data = drought_exp_clean) #dataframe
```

#### Model 4 - **total biomass as a function of water treatment and species**

```{r}
model4 <- lm(total_g ~ water_treatment + species_name, #formula
             data = drought_exp_clean) #dataframe
```

### a. Table of Models

```{r}
#compare the models
model_comparison <- modelsummary::modelsummary( # this function takes a list of models
  list( 
    "null" = model0, # "model name" = model object for all below 
    "model 1" = model1,
    "model 2" = model2,
    "model 3" = model3,
    "model 4" = model4))
model_comparison
```

**Table 1:** **Comparison of Regression Models.** This table compares five regression models: "null," "model 1," "model 2," "model 3," and "model 4." The columns represent the different models. The rows include the intercept, coefficients for the predictor variables (`sla`, `water_treatmentWell watered`, various `species_name` variables), and their standard errors in parentheses. Summary statistics provided for each model include the number of observations (Num.Obs.), R-squared (R2), adjusted R-squared (R2 Adj.), Akaike Information Criterion (AIC), Bayesian Information Criterion (BIC), log-likelihood (Log.Lik.), and root mean squared error (RMSE).

### b. Statistical Methods

To examine the influence of specific leaf area (mm^2^/g), watering level (well watered vs drought stressed, and species type on total biomass, I compared the five models outlined in the table above in order to determine which predictor variables best influence total biomass. I used the `model.sel` function from the `MuMIn` package which uses corrected (for smaller sample size) Akaike Information Criterion (AICc) to rank the provided models according to their AICc values. Model 4, which looks at total biomass as a function of water treatment and species, has the lowest AICc value (-156.2) and highest weight (0.772), indicating it is the best model among those compared. To evaluate linear model assumptions, I used the `plot` for each model to evaluate the diagnostic plots for residuals vs fitted values, the Normal Q-Q plot, the Scale-Location plot, and the residuals vs. leverage plot. These visual checks confirmed the models were homoscedastic, normally distributed, and linear.

### c. Visualization

```{r}
# creating new data frame of model predictions for plotting
model_preds_for_plotting <- model_preds %>% 
  rename(sla = x, # renaming columns to make this easier to use
         water_treatment = group,
         species_name = facet)

ggplot() +
  # underlying data
  geom_point(data = drought_exp_clean,
             aes(x = sla,
                 y = total_g,
                 color = water_treatment)) +
  # model prediction 95% CI ribbon
  geom_ribbon(data = model_preds_for_plotting,
              aes(x = sla, 
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = water_treatment),
              alpha = 0.2) +
  # model prediction lines
  geom_line(data = model_preds_for_plotting,
            aes(x = sla, 
                y = predicted,
                color = water_treatment)) +
  # cleaner theme
  theme_classic() +
  # creating different panels for species
  facet_wrap(~species_name) 
```
